---
title: "case_study_ride_share_data"
author: "Furugh"
date: "5/28/2022"
output:
  pdf_document: default
  html_document: default
---

## Setting Up The Analysis
note: setting up the R environment for analysis by loading the data, and required packages.
```{r loading pachages}

library(tidyverse)
library(lubridate)
library(ggplot2)

```
### Step 1: collect the data
```{r loading 4 files of data}

q1_2022<- read_csv("Desktop/Google Capstone/Case Study 1/JAn/202201-divvy-tripdata.csv")
q2_2022<- read_csv("Desktop/Google Capstone/Case Study 1/Feb/202202-divvy-tripdata.csv")
q3_2022<- read_csv("Desktop/Google Capstone/Case Study 1/March/202203-divvy-tripdata.csv")
q4_2022<- read_csv("Desktop/Google Capstone/Case Study 1/Apr/202204-divvy-tripdata 2.csv")
```

### STEP 2: WRANGLE DATA AND COMBINE INTO A SINGLE FILE
```{r}
q1_2022<- mutate(q1_2022, ride_id = as.character(ride_id)
                 ,rideable_type = as.character(rideable_type))
q2_2022<-mutate(q2_2022, ride_id = as.character(ride_id)
                ,rideable_type = as.character(rideable_type))
q2_2022<-mutate(q3_2022, ride_id = as.character(ride_id)
                ,rideable_type = as.character(rideable_type))
q2_2022<-mutate(q4_2022, ride_id = as.character(ride_id)
                ,rideable_type = as.character(rideable_type))
  
all_trips<- bind_rows(q1_2022,q2_2022, q3_2022, q4_2022)

all_trips <- all_trips %>% 
  select(-c(start_lat, start_lng, end_lat, end_lng))
```

### STEP 3: CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS

```{r For consistancy, changing some costumer type names. And adding date columns}
all_trips <-  all_trips %>% 
  mutate(member_casual = recode(member_casual
                                ,"Subscriber" = "member"
                                ,"Customer" = "casual"))

table(all_trips$member_casual)

all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

```{r Adding a ride lengh calculation }
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)

is.factor(all_trips$ride_length)
is.numeric(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

### Removing bad data

```{r Removing NAs and unrequired data }
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0 | all_trips$member_casual == "NA"),]
all_trips_v2 <- na.omit(all_trips_v2)
```

### STEP 4: CONDUCT DESCRIPTIVE ANALYSIS

```{r Descriptive analysis on ride_length all figures in seconds mean values and other statistics, and distribution of ride durtaion around mean}


all_trips_v2 %>% 
  group_by(day_of_week) %>% 
  summarise(mean_per_day = mean(ride_length))

sd(all_trips_v2$ride_length)
mean(all_trips_v2$ride_length)
median(all_trips_v2$ride_length)
range(all_trips_v2$ride_length)
max(all_trips_v2$ride_length) 
min(all_trips_v2$ride_length) 


hist(table(all_trips_v2$ride_length), breaks = 10, col = "gray",  xlab = "Histogram with Normal Curve",
     main = "Trips Duration Distribution") %>%
  abline(v = mean(all_trips_v2$ride_length) , col = "blue")
```

```{r Compareing members and casual users by simple statistics}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)

all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week,
                                  levels=c("Sunday", "Monday", "Tuesday", 
                                           "Wednesday", "Thursday", "Friday", "Saturday"))

  
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```



```{r analyze ridership data by type and weekday}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday)	
```

```{r  visualizing the number of rides by rider type}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")



```

```{r}
all_trips_v2 %>%
  group_by(member_casual) %>%
  summarise(number_of_rides = n()) %>%
  ggplot(aes(x = member_casual, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```

```{r  creatting a visualization for average duration}
all_trips_v2 %>% 
  group_by(member_casual, day_of_week) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  ggplot(aes(x = day_of_week, y = average_duration, fill = member_casual)) +
  geom_col(position =  "dodge")